---
title: "Project Milestone 4"
output: html_document
date: "2023-11-16"
---

**Scenario 1**\
**Jourdan Jerome, Kateryna Louchnikov, Alyssa Nguyen**

```{r , include=FALSE, message=FALSE, warning=FALSE}

#Set packages and import data

library(tidyverse)
library(dplyr)
library(plotly)  
library(kableExtra)

source_vax_CA <- read.csv("~/PHW251_group_project/data/ca_vax_rates_quarter.csv")
source_flu_LA <- read.csv("~/PHW251_group_project/data/sim_flu_LACounty_pop.csv")
source_flu_CA <- read.csv("~/PHW251_group_project/data/sim_flu_CA.csv")

```

```{r , include=FALSE, message=FALSE warning= FALSE}

##Clean and stack flu data sets
#Clean CA flu data

flu_CA <- source_flu_CA %>%
  mutate(county = str_replace(county," County",""),
         sex = str_to_title(sex),
         race_ethnicity = case_when(
           race_ethnicity == "1" ~ "White, Non-Hispanic",
           race_ethnicity == "2" ~ "Black, Non-Hispanic",
           race_ethnicity == "3" ~ "American Indian or Alaska Native, Non-Hispanic",
           race_ethnicity == "4" ~ "Asian, Non-Hispanic",
           race_ethnicity == "5" ~ "Native Hawaiian or Pacific Islander, Non-Hispanic",
           race_ethnicity == "6" ~ "Multiracial (two or more of above races), Non-Hispanic",
           race_ethnicity == "7" ~ "Hispanic (any race)",
           race_ethnicity == "9" ~ "Unknown"),
         report_date = as.Date(report_date, format = "%Y-%m-%d"),
         dt_diagnosis = as.Date(dt_diagnosis, format = "%Y-%m-%d"))

#Clean LA flu data to match CA flu data

flu_LA <- source_flu_LA %>%
  rename(new_infections = DX_NEW,
         new_recovered = RECOVERED_NEW,
         count_susceptible = SUSCEPTIBLE,
         current_infected = INFECTED_CURRENT,
         cumulative_infected = INFECTED_CUMULATIVE,
         cumulative_recovered = RECOVERED_CUMULATIVE,
         new_severe = SEVERE_NEW,
         cumulative_severe = SEVERE_CUMULATIVE,
         age_cat = AGE_CATEGORY,
         sex = SEX,
         race_ethnicity = RACE_ETH,
         report_date = DT_REPORT,
         dt_diagnosis = DT_DX,
         pop = POPULATION) %>%
  mutate(county = "Los Angeles",
         sex = str_to_title(sex),
         dt_diagnosis = as.Date(dt_diagnosis, format = "%d%b%Y"),
         time_int = NA_character_,
         new_unrecovered = NA_real_,
         cumulative_unrecovered = NA_real_) %>%
  select(county,
         time_int,
         new_infections,
         new_recovered,
         count_susceptible,
         current_infected,
         cumulative_infected,
         cumulative_recovered,
         new_unrecovered,
         cumulative_unrecovered,
         new_severe,
         cumulative_severe,
         age_cat,
         sex,
         race_ethnicity,
         report_date,
         dt_diagnosis,
         pop)

#Stack CA and LA flu data

flu_CA_LA <- rbind(flu_CA, flu_LA) %>%
  mutate(quarter = as.character(floor_date(dt_diagnosis, "quarter")))

## Restructure flu data so one demographic variable per row

#only one demographic value case per row 

flu_age <- flu_CA_LA %>%
  group_by(county,age_cat,quarter) %>%
  summarise(new_infections = sum(new_infections),
            new_severe = sum(new_severe)) %>%
  ungroup()  %>%
  mutate(demographic_category = "Age Group") %>%
  rename(demographic_value = age_cat)

flu_sex <- flu_CA_LA %>%
  group_by(county,sex,quarter) %>%
  summarise(new_infections = sum(new_infections),
            new_severe = sum(new_severe)) %>%
  ungroup() %>%
  mutate(demographic_category = "Gender") %>%
  rename(demographic_value = sex)

flu_race <- flu_CA_LA %>%
  group_by(county,race_ethnicity,quarter) %>%
  summarise(new_infections = sum(new_infections),
            new_severe = sum(new_severe)) %>%
  ungroup() %>%
  mutate(demographic_category = "Race/Ethnicity") %>%
  rename(demographic_value = race_ethnicity)

flu_cases <- rbind(flu_age, flu_sex, flu_race)

#only one demographic population group per row

pop_flu_age <- flu_CA_LA %>%
  distinct(county,age_cat,sex,race_ethnicity,pop) %>%
  group_by(county,age_cat) %>%
  summarise(pop = sum(pop)) %>%
  ungroup() %>%
  rename(demographic_value = age_cat) %>%
  mutate(demographic_category = "Age Group")

pop_flu_sex <- flu_CA_LA %>%
  distinct(county,age_cat,sex,race_ethnicity,pop) %>%
  group_by(county,sex) %>%
  summarise(pop = sum(pop)) %>%
  ungroup() %>%
  rename(demographic_value = sex) %>%
  mutate(demographic_category = "Gender")

pop_flu_race <- flu_CA_LA %>%
  distinct(county,age_cat,sex,race_ethnicity,pop) %>%
  group_by(county,race_ethnicity) %>%
  summarise(pop = sum(pop)) %>%
  ungroup() %>%
  rename(demographic_value = race_ethnicity) %>%
  mutate(demographic_category = "Race/Ethnicity")

flu_pop <- rbind(pop_flu_age, pop_flu_sex, pop_flu_race)

#Join both new_infection and pop to create final flu data table (flu)

flu <- flu_cases %>%
  full_join(flu_pop, by = c("county","demographic_value","demographic_category")) 

```

```{r , include=FALSE, message=FALSE, warning=FALSE}

#Clean COVID data

vax <- source_vax_CA %>%
  rename_with(~ tolower(gsub(".","_",.x,fixed=TRUE))) %>%
  rename(pop = estimated_population,
         county = county_name) %>%
  mutate(dt_admin = as.Date(dt_admin, format = "%Y-%m-%d"),
         demographic_value = case_when(
           demographic_value == "Under 5" ~ "0-17",
           demographic_value == "5-11" ~ "0-17",
           demographic_value == "12-17" ~ "0-17",
           demographic_value == "American Indian or Alaska Native" ~ "American Indian or Alaska Native, Non-Hispanic",
           demographic_value == "Asian" ~ "Asian, Non-Hispanic",
           demographic_value == "Black or African American" ~ "Black, Non-Hispanic",
           demographic_value == "Latino" ~ "Hispanic (any race)",
           demographic_value == "Multiracial" ~ "Multiracial (two or more of above races), Non-Hispanic",
           demographic_value == "Native Hawaiian or Other Pacific Islander" ~ "Native Hawaiian or Pacific Islander, Non-Hispanic",
           demographic_value == "White" ~ "White, Non-Hispanic",
           #  demographic_value == "Other Race" ~ "Unknown",
           #  demographic_value == "Unknown/undifferentiated" ~ "Unknown gender"
           TRUE ~ demographic_value)) %>%
  group_by(county,demographic_category,demographic_value,quarter) %>%
  summarise(pop=sum(pop),
            total_partial_vaccinated = sum(total_partial_vaccinated),
            cumulative_fully_vaccinated = sum(cumulative_fully_vaccinated),
            cumulative_at_least_one_dose = sum(cumulative_at_least_one_dose),
            cumulative_up_to_date_vax = sum(cumulative_up_to_date_vax)) %>%
  ungroup() %>%
  filter(quarter >= "2022-07-01" & quarter <= "2023-04-01",
         county != "Statewide")

```

```{r , include=FALSE, message=FALSE, warning=FALSE}

#Aggregate flu data

flu_counts_by_demo <- flu %>%
  group_by(demographic_category,demographic_value) %>%
  summarise(new_infections = sum(new_infections),
            new_severe = sum(new_severe)) %>%
  ungroup()

flu_pop_demo_state <- flu_pop %>%
  group_by(demographic_category,demographic_value) %>%
  summarise(pop = sum(pop)) %>%
  ungroup()

flu_pop_county <- flu_pop %>%
  filter(demographic_category == "Age Group") %>%
  group_by(county) %>%
  summarise(pop = sum(pop)) %>%
  ungroup()

flu_pop_ca <- flu_pop %>%
  filter(demographic_category == "Age Group") %>%
  summarise(pop = sum(pop)) %>%
  mutate(county = "California")

flu_rate_by_demo <- flu_counts_by_demo %>%
  left_join(flu_pop_demo_state, by = c("demographic_value","demographic_category")) %>%
  mutate(rate_infections = (new_infections/pop)*1000,
         rate_severe = (new_severe/pop)*1000) %>%
  ungroup()

flu_rate_by_race <- flu_rate_by_demo %>%
  filter(demographic_category == "Race/Ethnicity")

flu_rate_by_age <- flu_rate_by_demo %>%
  filter(demographic_category == "Age Group")

flu_rate_by_sex <- flu_rate_by_demo %>%
  filter(demographic_category == "Gender")

flu_rate_by_quarter <- flu %>%
  filter(demographic_category == "Gender") %>%
  group_by(quarter) %>%
  summarise(new_infections = sum(new_infections),
            new_severe = sum(new_severe)) %>%
  ungroup() %>%
  mutate(county = "California") %>%
  left_join(flu_pop_ca, by = "county") %>%
  mutate(rate_infections = round((new_infections/pop)*1000, digits=2),
         rate_severe =round((new_severe/pop)*1000, digits =2))

```

```{r , include=FALSE, message=FALSE, warning=FALSE}

#Aggregate COVID vax data

vax_count_by_demo <- vax %>%
  group_by(demographic_category,demographic_value) %>%
  summarise(total_partial_vaccinated = sum(total_partial_vaccinated, na.rm = TRUE),
            cumulative_at_least_one_dose = sum(cumulative_at_least_one_dose, na.rm = TRUE)) %>%
  ungroup()

vax_pop <- vax %>%
  distinct(county,demographic_category,demographic_value, pop) %>% 
  filter(demographic_value != "Unknown/undifferentiated", demographic_value != "Other Race", demographic_value != "Unknown")

vax_pop_demo_state <- vax_pop %>%
  group_by(demographic_category,demographic_value) %>%
  summarise(pop = sum(pop, na.rm = TRUE)) %>%
  ungroup()  

vax_pop_county <- vax_pop %>%
  filter(demographic_category == "Age Group") %>%
  group_by(county) %>%
  summarise(pop = sum(pop))

vax_pop_ca <- vax_pop_county %>%
  summarise(pop = sum(pop)) %>%
  mutate(county = "California")

vax_rate_by_demo <- vax_count_by_demo %>%
  left_join(vax_pop_demo_state, by = c("demographic_value","demographic_category")) %>%
  mutate(rate_total_partial = (total_partial_vaccinated/pop)*1000,
         rate_cum_one_dose = (cumulative_at_least_one_dose/pop)*1000) %>%
  ungroup() %>%
  drop_na()

vax_rate_by_race <- vax_rate_by_demo %>%
  filter(demographic_category == "Race/Ethnicity")

vax_rate_by_age <- vax_rate_by_demo %>%
  filter(demographic_category == "Age Group")

vax_rate_by_sex <- vax_rate_by_demo %>%
  filter(demographic_category == "Gender")

vax_rate_by_quarter <- vax %>%
  filter(demographic_category == "Gender") %>%
  group_by(quarter) %>%
  summarise(total_partial_vaccinated = sum(total_partial_vaccinated, na.rm = TRUE),
            cumulative_at_least_one_dose = sum(cumulative_at_least_one_dose, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(county = "California") %>%
  left_join(vax_pop_ca, by = "county")  %>%
  mutate(rate_total_partial = round((total_partial_vaccinated/pop)*1000, digits = 2),
         rate_cum_one_dose = round((cumulative_at_least_one_dose/pop)*1000, digits = 2))

```

### Joining Datasets

```{r, message=FALSE, warning=FALSE}

flu_vax <- flu %>%
  full_join(vax, by = c("county","demographic_value","demographic_category","quarter"))

str(flu_vax)
```

### Visualizations

**Table**: Rates of Severe Flu and COVID Vaccination by Demographic Group

```{r, message=FALSE, warning=FALSE, echo = TRUE}
flu_vax_rate_by_demo <- flu_vax %>%
  group_by(demographic_category,demographic_value) %>%
  summarise(new_severe = sum(new_severe, na.rm = TRUE),
            total_partial_vaccinated = sum(total_partial_vaccinated, na.rm = TRUE)) %>%
  ungroup() %>%
  left_join(flu_pop_demo_state, by = "demographic_value") %>%
  left_join(vax_pop_demo_state, by = "demographic_value") %>%
  mutate(rate_severe = round((new_severe/pop.x)*1000,2),
         rate_partial_vaccinated = round((total_partial_vaccinated/pop.y)*1000,2)) %>%
  select(demographic_value, rate_severe, rate_partial_vaccinated) %>%
  drop_na()

str(flu_vax_rate_by_demo)

kable(flu_vax_rate_by_demo,longtable = TRUE, booktabs = TRUE, col.names = c("Demographic Group", "Severe Flu", "COVID Vaccination"), caption = "Rates of Severe Flu and COVID Vaccination by Demographic Group, California") %>%
  kable_styling(full_width = FALSE) %>%
  kable_styling(position = "left") %>%
  kable_styling(font_size = 10) %>%
  add_header_above(c(" "=1,"Rate per 1,000 population"=2)) %>%
  pack_rows("Age Group",1,4)%>%
  pack_rows("Gender",5,6) %>%
  pack_rows("Race/Ethnicity",7,13) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  # add_footnote("Rates are per 1,000 population", notation = "none") %>%
  add_footnote("Based on diagnosis date from 9/24/2022 - 6/10/2023", notation = "none")
```

Interpretation: By age group, severe flu incidence increases with older age groups and vaccination is lowest in the youngest group (0-17 years). By race/ethnicity, severe flu incidence is highest within the White, non-Hispanic group and vaccination is lowest in the Multi-racial group followed by the White, non-Hispanic group.

**Scatterplot**: County Rates of Totally Partially Vaccinated for COVID-19 and Severe Flu

```{r, message=FALSE, warning=FALSE, echo = TRUE}

flu_vax_by_county <- flu_vax %>%
  filter(demographic_category == "Gender") %>%
  group_by(county) %>%
  summarise(new_severe = sum(new_severe, na.rm = TRUE),
            total_partial_vaccinated = sum(total_partial_vaccinated, na.rm = TRUE)) %>%
  ungroup() %>%
  left_join(flu_pop_county, by = "county") %>%
  left_join(vax_pop_county, by = "county") %>%
  mutate(rate_severe = round((new_severe/pop.x)*1000,2),
         rate_partial_vaccinated = round((total_partial_vaccinated/pop.y)*1000,2)) %>%
  filter(county != "Imperial")

str(flu_vax_by_county)

plot_ly(
  flu_vax_by_county,
  x = ~rate_partial_vaccinated,
  y = ~rate_severe,
  type = "scatter",
  mode = "markers",
  marker = list(color = "darkcyan"),  # Corrected 'colors' to 'marker'
  text = ~paste('County: ', county, '<br>COVID Vaccination: ', rate_partial_vaccinated, '<br>Severe Flu: ', rate_severe)
) %>%
  layout(
    title = "County Rates of Totally Partially Vaccinated for COVID-19 and Severe Flu",
    yaxis = list(title = "Severe Flu Rate per 1,000"),
    xaxis = list(title = "Total Partially Vaccinated for COVID-19 Rate per 1,000")
  )

```

Interpretation: There is not a strong correlation at the county-level between rates of severe flu and total partially vaccinated for COVID-19. Severe flu rates ranged from 7.1 to 15.8 per 1,000, and COVID-19 vaccination ranged from 114.34 to 452.69 per 1,000. The data point for Imperial County was excluded due to outliner value (COVID-19 vaccination rate is 1728.22 per 1,000 population).

**Faceted Subplot:** Rates for Severe Flu and COVID Partially Vaccinated by Race/Ethnicity in California

```{r, message=FALSE, warning=FALSE, echo = TRUE}
flu_vax_race <- full_join(flu_rate_by_race, vax_rate_by_race, by = "demographic_value")
str (flu_vax_race)

plot1 <- plot_ly(flu_vax_race,
                 x = ~demographic_value,
                 y = ~rate_severe,
                 name = "Severe Flu",
                 type = "bar",
                 marker = list(color = "darkcyan"))

plot2 <- plot_ly(flu_vax_race,
                 x = ~demographic_value,
                 y = ~rate_total_partial,
                 name = "COVID Partially Vaccinated",
                 type = "bar",
                 marker = list(color = "darkslateblue"))

subplot(plot1, plot2, nrows = 2, shareX = TRUE) %>%
  layout(
    xaxis = list(title = ""),
    yaxis1 = list(title = "Rate per 1,000"),
    yaxis2 = list(title = "Rate per 1,000"),
    title = "Rates for Severe Flu and COVID Partially Vaccinated by Race/Ethnicity in California"
  )
```

Interpretation:

**Bar Chart:**

```{r, message=FALSE, warning=FALSE. echo = TRUE}
flu_vax_quarter <- inner_join (vax_rate_by_quarter,flu_rate_by_quarter, by ="quarter")
str(flu_vax_quarter)


plot_ly(flu_vax_quarter, x = ~quarter, y = ~rate_total_partial, type = "bar", name = "Partial COVID Vaccination", marker = list(color = "darkcyan")) %>%
  add_trace(x = ~quarter, y = ~rate_infections, type = "bar", name = "Flu Diagnosed", marker = list(color = "darkslateblue")) %>%
  layout(
    title = "Rates for Partially COVID Vaccinated and Newly Flu Diagnosed by Quarter in California, 2022-2023",
    yaxis = list(title = "Rate per 1,000"),
    xaxis = list(title = "Quarters"),
    plot_bgcolor = "white"
  )

```

Interpretation: The chart above shows that rates of newly flu diagnosis per 1,000 individuals do not have a strong correlation with rates of partially COVID vaccination per 1,000 individuals across quarters from July 2022 to April 2023. The observed increase in flu diagnosis rates during the second quarter and the largest spike in the third quarter may be attributed to the flu season.
