---
title: "Project Milestone 3"
output: html_document
date: "2023-10-26"
editor_options: 
  markdown: 
    wrap: 72
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Import data

-   Set packages and import data

```{r}
library(tidyverse)
library(dplyr)

source_vax_CA <- read.csv("~/PHW251_group_project/data/ca_vax_rates_quarter.csv")
source_flu_LA <- read.csv("~/PHW251_group_project/data/sim_flu_LACounty_pop.csv")
source_flu_CA <- read.csv("~/PHW251_group_project/data/sim_flu_CA.csv")

#str(source_flu_CA)
#str(source_flu_LA)
#str(source_vax_CA)

#unique(source_flu_CA$race_ethnicity)
#unique(source_flu_LA$RACE_ETH)
```

### Clean and stack flu data sets

-   Clean CA flu data

    -   rewrite values in the sex column to lower case

    -   rewrite race/ethnicity categories according to the
        categorization as defined by California Department of Finance

    -   change data types for two columns

```{r}
flu_CA <- source_flu_CA %>%
  mutate(county = str_replace(county,"County",""),
         sex = str_to_title(sex),
         race_ethnicity = case_when(
           race_ethnicity == "1" ~ "White, Non-Hispanic",
           race_ethnicity == "2" ~ "Black, Non-Hispanic",
           race_ethnicity == "3" ~ "American Indian or Alaska Native, Non-Hispanic",
           race_ethnicity == "4" ~ "Asian, Non-Hispanic",
           race_ethnicity == "5" ~ "Native Hawaiian or Pacific Islander, Non-Hispanic",
           race_ethnicity == "6" ~ "Multiracial (two or more of above races), Non-Hispanic",
           race_ethnicity == "7" ~ "Hispanic (any race)",
           race_ethnicity == "9" ~ "Unknown"),
  report_date = as.Date(report_date, format = "%Y-%m-%d"),
  dt_diagnosis = as.Date(dt_diagnosis, format = "%Y-%m-%d"))

```

-   Clean LA flu data to match CA flu data
    -   renamed column names to match CA flu data
    -   created new columns to match CA flu data

```{r}

flu_LA <- source_flu_LA %>%
  rename(new_infections = DX_NEW,
         new_recovered = RECOVERED_NEW,
         count_susceptible = SUSCEPTIBLE,
         current_infected = INFECTED_CURRENT,
         cumulative_infected = INFECTED_CUMULATIVE,
         cumulative_recovered = RECOVERED_CUMULATIVE,
         new_severe = SEVERE_NEW,
         cumulative_severe = SEVERE_CUMULATIVE,
         age_cat = AGE_CATEGORY,
         sex = SEX,
         race_ethnicity = RACE_ETH,
         report_date = DT_REPORT,
         dt_diagnosis = DT_DX,
         pop = POPULATION) %>%
  mutate(county = "Los Angeles",
         sex = str_to_title(sex),
         dt_diagnosis = as.Date(dt_diagnosis, format = "%d%b%Y"),
         time_int = NA_character_,
         new_unrecovered = NA_real_,
         cumulative_unrecovered = NA_real_) %>%
  select(county,
         time_int,
         new_infections,
         new_recovered,
         count_susceptible,
         current_infected,
         cumulative_infected,
         cumulative_recovered,
         new_unrecovered,
         cumulative_unrecovered,
         new_severe,
         cumulative_severe,
         age_cat,
         sex,
         race_ethnicity,
         report_date,
         dt_diagnosis,
         pop)

```

-   Stack CA and LA flu data, and restructure to match vax data

```{r}

flu_CA_LA <- rbind(flu_CA, flu_LA)

#colSums(is.na(flu_CA_LA))
```

-   Restructure flu data so one demographic variable per row

```{r}
flu_age <- flu_CA_LA %>%
   group_by(county,age_cat,dt_diagnosis) %>%
   summarise(new_infections = sum(new_infections),
             new_severe = sum(new_severe)) %>%
   ungroup()  %>%
   mutate(demographic_category = "Age Group") %>%
   rename(demographic_value = age_cat)

flu_sex <- flu_CA_LA %>%
   group_by(county,sex,dt_diagnosis) %>%
   summarise(new_infections = sum(new_infections),
             new_severe = sum(new_severe)) %>%
   ungroup() %>%
   mutate(demographic_category = "Gender") %>%
   rename(demographic_value = sex)

 flu_race <- flu_CA_LA %>%
   group_by(county,race_ethnicity,dt_diagnosis) %>%
   summarise(new_infections = sum(new_infections),
             new_severe = sum(new_severe)) %>%
   ungroup() %>%
   mutate(demographic_category = "Race/Ethnicity") %>%
   rename(demographic_value = race_ethnicity)

flu_cases <- rbind(flu_age, flu_sex, flu_race)

# var <- c("age_cat", "sex", "race_ethnicity")
# value <- c("Age Group", "Gender", "Race/Ethnicity")
#
# for (i in 1:length(var)) {
#   flu_var[i] <- flu_CA_LA %>%
#     group_by(county,var[i],dt_diagnosis) %>%
#     summarise(new_infections = sum(new_infections)) %>%
#     ungroup() %>%
#     mutate(demographic_category = value[i]) %>%
#     rename(demographic_value = var[i])
# }
```

-   Restructure flu pop column so only one demographic variable per
    column

```{r}
 pop_flu_age <- flu_CA_LA %>%
   distinct(county,age_cat,sex,race_ethnicity,pop) %>%
   group_by(county,age_cat) %>%
   summarise(pop = sum(pop)) %>%
   ungroup() %>%
   rename(demographic_value = age_cat) %>%
   mutate(demographic_category = "Age Group")

 pop_flu_sex <- flu_CA_LA %>%
   distinct(county,age_cat,sex,race_ethnicity,pop) %>%
   group_by(county,sex) %>%
   summarise(pop = sum(pop)) %>%
   ungroup() %>%
   rename(demographic_value = sex) %>%
   mutate(demographic_category = "Gender")

 pop_flu_race <- flu_CA_LA %>%
   distinct(county,age_cat,sex,race_ethnicity,pop) %>%
   group_by(county,race_ethnicity) %>%
   summarise(pop = sum(pop)) %>%
   ungroup() %>%
   rename(demographic_value = race_ethnicity) %>%
   mutate(demographic_category = "Race/Ethnicity")

 flu_pop <- rbind(pop_flu_age, pop_flu_sex, pop_flu_race)
```

-   Join both new_infection and pop to create final flu data table (flu)
    -   Create new variable to capture month of diagnosis

```{r}

flu <- flu_cases %>%
   full_join(flu_pop, by = c("county","demographic_value","demographic_category")) %>%
   mutate(year_month = format(as.Date(dt_diagnosis), "%Y-%m"))

```

### Clean COVID data

-   Recode age categories
-   Change data types for several columns
-   Rename columns to match flu data
-   Find overlapping period between two data sets - from 2022-09-24 to
    2023-06-10

```{r}

vax <- source_vax_CA %>%
  rename_with(~ tolower(gsub(".","_",.x,fixed=TRUE))) %>%
  rename(pop = estimated_population,
         county = county_name) %>%
  mutate(dt_admin = as.Date(dt_admin, format = "%Y-%m-%d"),
    demographic_value = case_when(
          demographic_value == "Under 5" ~ "0-17",
          demographic_value == "5-11" ~ "0-17",
          demographic_value == "12-17" ~ "0-17",
          demographic_value == "American Indian or Alaska Native" ~ "American Indian or Alaska Native, Non-Hispanic",
          demographic_value == "Asian" ~ "Asian, Non-Hispanic",
          demographic_value == "Black or African American" ~ "Black, Non-Hispanic",
          demographic_value == "Latino" ~ "Hispanic (any race)",
          demographic_value == "Multiracial" ~ "Multiracial (two or more of above races), Non-Hispanic",
          demographic_value == "Native Hawaiian or Other Pacific Islander" ~ "Native Hawaiian or Pacific Islander, Non-Hispanic",
          demographic_value == "White" ~ "White, Non-Hispanic",
                              #  demographic_value == "Other Race" ~ "Unknown",
                              #  demographic_value == "Unknown/undifferentiated" ~ "Unknown gender"
          TRUE ~ demographic_value)) %>%
  group_by(county,demographic_category,demographic_value,dt_admin) %>%
  summarise(pop=sum(pop),
          total_partial_vaccinated = sum(total_partial_vaccinated),
          cumulative_fully_vaccinated = sum(cumulative_fully_vaccinated),
          cumulative_at_least_one_dose = sum(cumulative_at_least_one_dose),
          cumulative_up_to_date_vax = sum(cumulative_up_to_date_vax)) %>%
  ungroup() %>%
  filter(dt_admin >= "2022-09-24" & dt_admin <= "2023-06-10") %>%
  mutate(year_month = format(as.Date(dt_admin), "%Y-%m"))

#min(vax$dt_admin)
#max(vax$dt_admin)

#min(flu$dt_diagnosis)
#max(flu$dt_diagnosis)

#unique(flu_CA_LA$age_cat)
#unique(vax$demographic_category)
#unique(source_vax_CA$Demographic.Value)

```

### Aggregate flu data

-   Create new variables needed for analysis

    -   rate_infections

    -   rate_severe

-   New variables are created based on existing columns: new_infections
    & new_severe

-   Create data sets that we want to focus on later

    -   flu_rate_by_demo - to see flu rate for certain demographic

    -   flu_counts_by_month - to track flu rates over time

```{r}

flu_counts_by_demo <- flu %>%
  group_by(demographic_category,demographic_value) %>%
  summarise(new_infections = sum(new_infections),
            new_severe = sum(new_severe)) %>%
  ungroup()

flu_pop_demo_state <- flu_pop %>%
  group_by(demographic_category,demographic_value) %>%
  summarise(pop = sum(pop)) %>%
  ungroup()

flu_pop_county <- flu_pop %>%
  group_by(county) %>%
  summarise(pop = sum(pop)) %>%
  ungroup()

flu_rate_by_demo <- flu_counts_by_demo %>%
  left_join(flu_pop_demo_state, by = c("demographic_value","demographic_category")) %>%
  mutate(rate_infections = (new_infections/pop)*1000,
         rate_severe = (new_severe/pop)*1000) %>%
  ungroup()

flu_rate_by_race <- flu_rate_by_demo %>%
  filter(demographic_category == "Race/Ethnicity")

flu_rate_by_age <- flu_rate_by_demo %>%
  filter(demographic_category == "Age Group")

flu_rate_by_sex <- flu_rate_by_demo %>%
  filter(demographic_category == "Gender")

flu_counts_by_month <- flu %>%
  group_by(year_month) %>%
  summarise(new_infections = sum(new_infections),
            new_severe = sum(new_severe)) %>%
  ungroup()

```

### Aggregate COVID vax data

-   Create new variables needed for analysis

    -   rate_total_partial

    -   rate_cum_one_dose

-   Handle missing data in pop column

-   New variables are created based on existing columns:
    total_partial_vaccinated & cumulative_at_least_one_dose

-   Create data sets that we want to focus on later

    -   vax_rate_by_demo - to see vaccination rates for certain
        demographic

    -   vax_counts_by_month - to track vaccination rates over time

```{r}
vax_count_by_demo <- vax %>%
  group_by(demographic_category,demographic_value) %>%
  summarise(total_partial_vaccinated = sum(total_partial_vaccinated, na.rm = TRUE),
            cumulative_at_least_one_dose = sum(cumulative_at_least_one_dose, na.rm = TRUE)) %>%
  ungroup()

vax_pop_demo_state <- vax %>%
  group_by(demographic_category,demographic_value) %>%
  summarise(pop = sum(pop, na.rm = TRUE)) %>%
  ungroup()  %>% 
  filter(demographic_value != "Unknown/undifferentiated", demographic_value != "Other Race", demographic_value != "Unknown")

vax_rate_by_demo <- vax_count_by_demo %>%
  left_join(vax_pop_demo_state, by = c("demographic_value","demographic_category")) %>%
  mutate(rate_total_partial = (total_partial_vaccinated/pop)*1000,
         rate_cum_one_dose = (cumulative_at_least_one_dose/pop)*1000) %>%
  ungroup() %>%
  drop_na()

vax_counts_by_month <- vax %>%
  group_by(year_month) %>%
  summarise(total_partial_vaccinated = sum(total_partial_vaccinated, na.rm = TRUE),
            cumulative_at_least_one_dose = sum(cumulative_at_least_one_dose, na.rm = TRUE)) %>%
  ungroup()

```

### Data Dictionary for Flu Data

-   Data dictionary based on clean data set

|                       |            |                                                                               |                                                                                                                                                                                                                                                                                                                      |       |
|--------|--------|--------|---------------------------------------|--------|
| **Column Name**       | Format     | Definition                                                                    | Values                                                                                                                                                                                                                                                                                                               | Notes |
| county                | Plain text | County of residence                                                           | California Counties                                                                                                                                                                                                                                                                                                  |       |
| demogr aphic_value    | Plain text | Age, Race /Ethnicity, Gender                                                  | 0-17, 18-49, 50-64, 65+; MALE, FEMALE; "White, Non -Hispanic", "Black, Non -Hispanic", "American Indian or Alaska Native, Non -Hispanic", "Asian, Non -Hispanic", "Native Hawaiian or Pacific Islander, Non -Hispanic", " Multiracial (two or more of above races), Non -Hispanic", "Hispanic (any race)", "Unknown) |       |
| d t_diagnosis         | Date       | Date of Diagnosis                                                             | 2022-09-26: 2023-06-10                                                                                                                                                                                                                                                                                               |       |
| new \_infections      | Numeric    | Number of newly diagnosed individuals                                         | 0-95742                                                                                                                                                                                                                                                                                                              |       |
| new_severe            | Numeric    | Newly identified individuals having severe disease requiring hosp italization | 0-2488                                                                                                                                                                                                                                                                                                               |       |
| demograph ic_category | Plain text | Age, Race /Ethnicity, Gender                                                  | Race /Ethnicity, Age Group, Gender                                                                                                                                                                                                                                                                                   |       |
| pop                   | Numeric    | Total population of demographic categories in county                          | 0-5153156                                                                                                                                                                                                                                                                                                            |       |
| year_month            | Plain text | Year and month of diagnosis                                                   | 2022 -09:2023-06                                                                                                                                                                                                                                                                                                     |       |

### Descriptive Statistics

-   Data type Description One or more tables with descriptive statistics
    for 4 data elements

```{r}

#Descriptive stats for count of new infections by race
summary(flu_rate_by_race$rate_severe)
#Descriptive stats for count of new severe infections by race
summary(flu_rate_by_race$rate_infections)

#Descriptive stats for count of new infections by age group
summary(flu_rate_by_age$rate_severe)
#Descriptive stats for count of new severe infections by age group
summary(flu_rate_by_age$rate_infections)

#Descriptive stats for count of new infections by month
summary(flu_counts_by_month$new_infections)
#Descriptive stats for count of new severe infections by month
summary(flu_counts_by_month$new_severe)

summary(vax_rate_by_demo$rate_total_partial)
summary(vax_rate_by_demo$rate_cum_one_dose)

summary(vax_counts_by_month$total_partial_vaccinated)
summary(vax_counts_by_month$cumulative_at_least_one_dose)
```

-   Html output that is professionally prepared for presentation

    -   Only the necessary information is outputted (you should
        suppress, for example, entire data frame outputs)
